---
title: "PSTAT 126 Project"
author: "Ana Caklovic and Sukanya Joshi"
date: "11/16/2018"
output:
  word_document: default
  pdf_document: default
---

```{r}
#student <- read.csv2(file.choose())
student <- read.csv("student-mat(1).csv", header = TRUE)
```

create new, smaller data frames to make it easier to plot and create scatterplot matrices to check correlations 
```{r}
newstudent1 = data.frame(student$school, student$sex, student$age, student$address, student$famsize, student$Pstatus)
newstudent2 = data.frame(student$Medu, student$Fedu, student$Mjob, student$Fjob, student$reason, student$guardian)
newstudent3 = data.frame(student$traveltime, student$studytime, student$failures, student$schoolsup, student$famsup, student$paid)
newstudent4 = data.frame(student$activities, student$nursery, student$higher, student$internet, student$romantic, student$famrel, student$freetime)
newstudent5 = data.frame(student$goout, student$Dalc, student$Walc, student$health)
newstudent6 = data.frame(student$absences, student$G1, student$G2, student$G3)

#scatter1 = pairs(newstudent1)
#scatter2 = pairs(newstudent2)
#scatter3 = pairs(newstudent3)
#scatter4 = pairs(newstudent4)
#scatter5 = pairs(newstudent5)
scatter6 = pairs(newstudent6)
```

DIAGNOSTICS (OLD MODEL):
```{r}
#par(mfrow =c(1,3))
absences = student$absences
finalgrade = student$G3
plot(absences, finalgrade, xlab="Number of Absences", ylab = "Final Grade (1-20)", main = "Final Grade vs. Number of Absences")

student.fit = lm(finalgrade ~ absences)
abline(student.fit) # intercept = 10.30327, slope: 0.01961
# can see our plot isn't linear, will need some kind of transformation

# Res vs. Fit:
x = student$absences
y = student$G3
xbar = mean(x)
ybar = mean(y)
yhat = fitted(student.fit)
e = y - yhat
plot(yhat, e, xlab = 'Fitted Values', ylab = 'Residual', main = 'Residual vs Fit')
abline(h = 0, lty = 2)
# clustering on the left (clear funnel effect)

#Normal Q-Q plot
student.res = resid(student.fit)
qqnorm(student.res)
qqline(student.res)
# very skewed data, especially on the left tail 
shapiro.test(student.res)
# p-value = 4.283e-12, reject H0: data is normal --> know it has normality problem 

summary(student.fit, data = "student")$r.squared
hist(e, xlab = 'Residuals', main = 'Histogram of Residuals')
```

TRANSFORMATIONS:
--> seem to have problems with everything (linearity, equal variances, normality) so take log of both X and Y
```{r message=FALSE, warning=FALSE}
x = student$absences
y = student$G3
x.new = log(x)
#x.new
y.new = log(y)
#y.new
#stud.fit.new = lm(y.new ~ x.new)
# error due to log = "-Inf" bc can't do log of 0, replace "-Inf" with "1"

library(plyr)
x.new = mapvalues(x.new, from = "-Inf", to = "1")
x.new = as.numeric(as.character(x.new))
y.new = mapvalues(y.new, from = "-Inf", to = "1")
y.new = as.numeric(as.character(y.new))
stud.fit.new = lm(y.new ~ x.new)
```

DIAGNOSTICS (NEW MODEL):
```{r}
# res vs fit:
yhat.new = fitted(stud.fit.new)
e.new = y.new - yhat.new
plot(yhat.new, e.new, xlab = 'Fitted Values', ylab = 'Residual', main = 'Residual vs Fit')
abline(h = 0, lty = 2)

# Q-Q plot:
qqnorm(e.new)
qqline(e.new)

# R^2
summary(stud.fit.new, data = "student")$r.squared
```

BOX COX TRANSFORMATION:
```{r message=FALSE, warning=FALSE}
library(plyr)
library(SemiPar)
library(MASS)
attach(student)
y2 = student$G3
y2 = mapvalues(y2, from = "0", to = "1")
y2 = as.numeric(as.character(y2))
bc = boxcox(y2 ~ x.new)

lambda = bc$x # lambda values
lik = bc$y # likelihood values for SSE
bc.df = cbind(lambda, lik)
sorted_bc = bc.df[order(-lik)] # values are sorted to identify the lambda value for the maximum log likelihood for obtaining minimum SSE
head(sorted_bc, n=10)

y_lambda = y2^(1.191919)
fit.new = lm(y_lambda ~ x.new)
summary(fit.new)$r.squared

yhat.new = fitted(fit.new)
e.new = y_lambda - yhat.new

# new scatterplot:
plot(x.new, y_lambda, xlab = 'ln(absences)', ylab = '(final grade)^(1.19)', main = 'Scatterplot of (final grade)^(1.19) \n vs. ln(absences)')
abline(fit.new)

# res vs. fit
plot(yhat.new, e.new, xlab = 'Fitted Values', ylab = 'Residual', main = 'Residual vs Fitted Values')
abline(h = 0, lty = 2)
hist(e.new, xlab = 'Residuals', main = 'Histogram of Residuals')

#normal Q-Q plot
qqnorm(e.new)
qqline(e.new)

# Box Cox in one line:
#boxcox(y2 ~ absences, lambda = seq(0,1,0.1))
```


SIGNIFICANCE OF OTHER PREDICTORS:

```{r}
### AIC - choosing bw our categorical predictors: (TEST 1)
mod0 = lm(finalgrade ~ 1)
mod.upper = lm(finalgrade ~ student$traveltime + student$studytime + student$failures + student$activities + student$freetime)
step(mod0, scope = list(mod0, upper = mod.upper))
```
The AIC test suggest that our best model would include the failures and traveltime categorical predictors.

Let's check a few more predictors: (focus on family life)
```{r}
# TEST 2
mod02 = lm(finalgrade ~ 1)
mod.upper2 = lm(finalgrade ~ student$Medu + student$Fedu + student$Mjob + student$Fjob + student$famsup + student$famrel)
step(mod02, scope = list(mod02, upper = mod.upper2))
```
The suggested model includes the Medu (mother's education) and famsup (family educational support) predictors 

Now, let's focus on predictors that look at social life:
```{r}
# TEST 3  
mod03 = lm(finalgrade ~ 1)
mod.upper3 = lm(finalgrade ~ student$goout + student$Dalc + student$Walc + student$romantic + student$activities + student$freetime)
step(mod03, scope = list(mod03, upper = mod.upper3))
```
We can see the best model includes the predictors goout and romantic. 

```{r}
summary(lm(finalgrade ~ student$absences + student$traveltime))
summary(lm(finalgrade ~ student$absences + student$failures))
```


INTERACTIONS
Now, we want to check if any of these predictors interact with absences:

```{r}
### AIC Model 1: failures and traveltime
mod.reduced = lm(finalgrade ~ absences + student$failures + student$traveltime)
mod.full = lm(finalgrade ~ absences + student$failures + student$traveltime + absences*student$failures + absences*student$traveltime)
anova(mod.reduced, mod.full)
# p-value = 0.006 < 0.05 --> can reject null
```


```{r}
### AIC Model 2: Medu and famsup
mod.reduced = lm(finalgrade ~ absences + student$Medu + student$famsup)
mod.full = lm(finalgrade ~ absences + + student$Medu + student$famsup + absences*student$Medu + absences*student$famsup)
anova(mod.reduced, mod.full)
# p-value = 0.11 > 0.05 --> fail to reject H0 (can't use full)

### AIC Model 3: goout and romantic
mod.reduced = lm(finalgrade ~ absences + student$goout + student$romantic)
mod.full = lm(finalgrade ~ absences + student$goout + student$romantic + absences*student$goout + absences*student$romantic)
anova(mod.reduced, mod.full)
# p-value = 0.4097 > 0.05 --> fail to reject H0 (can't use full)
```

------Research questions we can test:

Question 2) T-tests for testing the significance of failures and traveltime
```{r}
# failures
#summary(lm(finalgrade ~ student$absences + student$failures)) (???)

summary(lm(finalgrade ~ student$failures))
```


```{r}
# traveltime 
#summary(lm(finalgrade ~ student$absences + student$traveltime)) (???)

summary(lm(finalgrade ~ student$traveltime))
```




1) For every amount of absences, is there a difference in the mean effect for the other two predictors? We need to test the null hypothesis: H0: B2 = B3 = B12 = B13 = 0 vs H1: at least one of these slope parameters is not 0.
```{r}
# using Model 1 (failures and traveltime)
mod.reduced = lm(finalgrade ~ absences)
mod.full = lm(finalgrade ~ absences + student$failures + student$traveltime + absences*student$failures + absences*student$traveltime)
anova(mod.reduced, mod.full)

# there's a significant difference in all the terms 
```

2) Does the effect of absence on the other predictor’s effectiveness depend
on that predictor? We need to test the null hypothesis H0: B12 = B13 = 0 vs H1: at least one of these slope parameters is not 0.
```{r}
# same code as seen above, when testing models for interactions with absences
### AIC Model 1: failures and traveltime
mod.reduced = lm(finalgrade ~ absences + student$failures + student$traveltime)
mod.full = lm(finalgrade ~ absences + student$failures + student$traveltime + absences*student$failures + absences*student$traveltime)
anova(mod.reduced, mod.full)
# p-value = 0.006 < 0.05 --> can reject null
# at least one of the interaction parameters is not zero
```

-> Now let us see which interaction term is significant:
```{r}
# Testing significance of absences*student$failures:
mod.reduced = lm(finalgrade ~ absences + student$failures + student$traveltime + absences*student$traveltime)
mod.full = lm(finalgrade ~ absences + student$failures + student$traveltime + absences*student$failures + absences*student$traveltime)
anova(mod.reduced, mod.full)

```
```{r}
# Testing significance of absences*student$traveltime:
mod.reduced = lm(finalgrade ~ absences + student$failures + student$traveltime + absences*student$failures)
mod.full = lm(finalgrade ~ absences + student$failures + student$traveltime + absences*student$failures + absences*student$traveltime)
anova(mod.reduced, mod.full)
# the interaction term that is significant is: absences*student$failures
# so the failures predictor has an interaction with absences; the traveltime predictor does not 
```



```{r}
# interaction between failures and traveltime?
mod.reduced = lm(finalgrade ~ absences + student$failures + student$traveltime + absences*student$failures)
mod.full = lm(finalgrade ~ absences + student$failures + student$traveltime + absences*student$failures + student$failures*student$traveltime)
anova(mod.reduced, mod.full)
```

Now, let us test the new model to ensure the predictors and interaction terms are all significant:
```{r}
#### AIC of new model:#####
mod0 = lm(finalgrade ~ absences)
mod.upper = lm(finalgrade ~ absences + student$failures + student$traveltime + absences*student$failures + absences*student$traveltime)
step(mod0, scope = list(mod0, upper = mod.upper))
```
FINAL MODEL: absences + failures + traveltime + absences x failures + failures x traveltime


-------------------------------------------------------------
Checking (running diagnostics on) the New/Final Model:

```{r}
### Scatterplot matrix:
finalstudent = data.frame(student$absences, student$failures, student$traveltime, absences*student$failures, student$G3)
pairs(finalstudent)
# just to take a closer look at correlations:
finalstudent2 = data.frame(student$absences, absences*student$failures, student$G3)
pairs(finalstudent2)

### FINAL MODEL:
finalmod = lm(finalgrade ~ absences + failures + traveltime + absences*failures + traveltime*failures, data= student)

### Residual vs. predictor:

# Residual vs. Absences (# other plots are the same since this is the only numerical vector)
x2 = student$absences
y = student$G3
xbar = mean(x2)
ybar = mean(y)
yhat = fitted(finalmod) 
e = y - yhat
plot(x2, e, xlab = 'Predictor (absences)', ylab = 'Residual', main = 'Residual vs Predictor (absences)')
abline(h = 0, lty = 2)
# clustering on the left (clear the fanning effect)


### Normal Q-Q plot
student.res = resid(finalmod)
finalstudent.res = resid(finalmod)
qqnorm(finalstudent.res) 
qqline(finalstudent.res)
# normality issue (diverges on the tails)

```

TRANSFORMATIONS ON FINAL MODEL:

```{r}
y = student$G3
x.new = log(student$absences) # can I just take the log of the numerical predictor???
#x.new
y.new = log(y)
#y.new
#stud.fit.new = lm(y.new ~ x.new)
# error due to log = "-Inf" bc can't do log of 0, replace "-Inf" with "1"

library(plyr)
x.new = mapvalues(x.new, from = "-Inf", to = "1")
x.new = as.numeric(as.character(x.new))
y.new = mapvalues(y.new, from = "-Inf", to = "1")
y.new = as.numeric(as.character(y.new))
stud.fit2.new = lm(y.new ~ x.new + failures + traveltime + absences*failures, data= student)

### DIAGNOTICS OF FINAL MODEL: 

# res vs predictor:
yhat.new = fitted(stud.fit2.new)
e.new = y.new - yhat.new
plot(x.new, e.new, xlab = 'Predictor', ylab = 'Residual', main = 'Residual vs Predictor')
abline(h = 0, lty = 2)

# Q-Q plot:
qqnorm(e.new)
qqline(e.new)
```

BOX COX TRANSFORMATION: 

```{r message=FALSE, warning=FALSE}
library(plyr)
library(SemiPar)
library(MASS)
attach(student)
y2 = student$G3
y2 = mapvalues(y2, from = "0", to = "1")
y2 = as.numeric(as.character(y2))
bc2 = boxcox(y2 ~ x.new + failures + traveltime + absences*failures, data= student)

lambda = bc2$x # lambda values
lik = bc2$y # likelihood values for SSE
bc.df = cbind(lambda, lik)
sorted_bc = bc.df[order(-lik)] # values are sorted to identify the lambda value for the maximum log likelihood for obtaining minimum SSE
head(sorted_bc, n=10)

y_lambda = y2^(1.151515)
fit.new = lm(y_lambda ~ x.new + failures + traveltime + absences*failures, data= student)
summary(fit.new)$r.squared

yhat.new = fitted(fit.new)
e.new = y_lambda - yhat.new

# new scatterplot:
plot(x.new, y_lambda, xlab = 'ln(absences)', ylab = '(final grade)^(1.15)', main = 'Scatterplot of (final grade)^(1.19) \n vs. ln(absences)')
abline(fit.new, col = 2)

# res vs. predictor
plot(x.new, e.new, xlab = 'Predictor (absences)', ylab = 'Residual', main = 'Residual vs Predictor (absences)')
abline(h = 0, lty = 2)
hist(e.new, xlab = 'Residuals', main = 'Histogram of Residuals')

#normal Q-Q plot
qqnorm(e.new)
qqline(e.new)

hist(e.new, xlab = 'Residuals', main = 'Histogram of Residuals (Final Model)')
```

